Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-05-29 Manuscript/Figures/figure_legends.Rmd
edits to figure legends

**Figure 2. Model Construction.** An illustration of the proposed model construction given cell line specific Hi-C matrices. The contact matrix above is a visualization of the interconnectivity of chromosome 14 at 100 kb resolution, for the GM12878 cell line, on the intervals from 20 mb to 100 mb [@durand2016juicer]. The boundary coordinates of topologically association domains (TADs) were identified using the Arrowhead algorithm. To account for the uncertainty in TAD boundary detection, each boundary point was flanked on both sides by resolution-specific genomic regions (bins). The response vector $Y$, used for classification, was determined by whether a genomic bin overlapped with a flanked boundary region ($Y=1$) or not ($Y=0$). The spatial relationship between positional coordinates of various functional genomic annotations, obtained from ENCODE, together with the genomic bins were then used to define the feature space of the models represented by the data matrix $X = \{X_1, X_2, ..., X_p\}$, where $p$ was the complete set of functional annotations.

**Figure 3. The three predictor types.** Three separate types of predictors were considered when quantifying the association between TAD boundaries and genomic annotations. Each predictor type was used to build the feature space of specific models in the ensemble framework. The _overlap count_ (OC) predictors consider the total number of genomic features that overlapped with each genomic bin. The _overlap percent_ (OP) predictors were calculated by dividing the sum of all feature widths within each specific bin by the total bin width. The _distance_ ($log_{2}$ transformed) predictors were calculated by measuring the distance (in bases) from the center of each genomic bin to the center of the nearest genomic feature.

**Figure 4. Comparing ensemble system specific model performances for GM12878.** Performances of random forest classification models built on each of the 3 predictor types: OC (A), OP (B), and Distance (C). Within each predictor type, and across resolution, MCCs metrics were recorded for models subsequently implemented on re-sampled data using the 4 re-sampling techniques: None (red), ROS (green), RUS (blue), and SMOTE (purple). Performances were aggregated using means across each chromosome, with error bars representing 1 standard deviation away from the mean.

**Figure 5. Predictive importance is outweighed by transcription factor binding sites.** Stacked barplots illustrating the proportion of total predictive importance in models using optimal data level characteristics determined by the ensemble framework, for each resolution on GM12878. Proportions are separated by the set of annotations included in the feature space including chromatin state segmentation, histone modification, and transcription factor binding sites.

>>>>

**Figure 2. Model Construction.** An illustration of the proposed model construction given cell line specific Hi-C data. Above is an example of a subsection of the contact matrix for chromosome 14 (at megabase scale) on GM12878 [@durand2016juicer]. The contact matrix is a visualization of the interconnectivity of the linear genome at 100 kb resolution. The boundary coordinates of TADs were identified using the Arrowhead algorithm. To account for the uncertainty in TAD boundary detection, each boundary point was flanked on both sides by resolution-specific genomic regions (bins). The response vector $Y$, used for classification, was determined by whether a genomic bin overlapped with a flanked boundary region ($Y=1$) or not ($Y=0$). The spatial relationship between positional coordinates of various functional genomic annotations, obtained from the UCSC Genome Browser Database, together with the genomic bins were then used to define the feature space of the models represented by the data matrix $X = \{X_1, X_2, ..., X_p\}$, where $p$ was the complete set of functional annotations.

**Figure 3. Predictor types.** Three separate types of predictors were considered when quantifying the association between TAD boundaries and genomic annotations. Each predictor type was used to build the feature space of specific models in the ensemble framework. (A) _Overlap Counts_ (OC) consider the total number of genomic features that overlapped with each genomic bin. (B) _Overlap Percent_ (OP) was calculated by dividing the sum of all feature widths within each specific bin by the total bin width. (C) _Distance_ ($log_{2}$ transformed) was calculated by measuring the distance (in bases) from the center of each genomic bin to the center of the nearest region defining a genomic annotation.

**Figure 4. SMOTE re-sampled models built on distance-type predictors yielded optimal MCCs across resolutions for GM12878.** Performances of random forest classification models built on each of the 3 predictor types: (A) OC, (B) OP, and (C) Distance. Within each predictor type, and across resolution, training data was re-sampled using different techniques including: None (red), ROS (green), RUS (blue), and SMOTE (purple). Performances were aggregated using means across each chromosome, with error bars representing 1 standard deviation away from the mean.

**Figure 5. The proportional importance of optimally built predictive models was outweighed by transcription factor binding sites across resolutions for GM12878.** Stacked barplots illustrate the proportion of total predictive importance in models using optimal data level characteristics determined by the ensemble framework (SMOTE re-sampling; with distance-type predictors), for each resolution on GM12878. Proportions are grouped by the class of cell line specific annotations in the feature space including chromatin state segmentations (CSS), histone modifications (HM), and transcription factor binding sites (TFBS).

### 2019-05-29 Manuscript/Supplementary/supplementary_legends.Rmd
edits to figure legends

**Supplementary Figure 1. GM12878 distance densities.** (A-D) Distances (in millions of basepairs) from the center of each genomic bin to the nearest annotation peak, for binwidths (resolutions) of 10 kb, 25 kb, 50 kb, and 100 kb. (E-H) $Log_{2}$ transformed distances for binwidths (resolutions) of 10 kb, 25 kb, 50 kb, and 100 kb.

**Supplementary Figure 2. K562 distance densities.** (A-D) Distances (in millions of basepairs) from the center of each genomic bin to the nearest annotation peak, for binwidths (resolutions) of 10 kb, 25 kb, 50 kb, and 100 kb. (E-H) $Log_{2}$ transformed distances for binwidths (resolutions) of 10 kb, 25 kb, 50 kb, and 100 kb.

@@

**Supplementary Figure 3. Comparing ensemble system specific model performances for K562.** Performances of random forest classification models built on each of the 3 predictor types: OC (A), OP (B), and Distance (C). Within each predictor type, and across resolution, MCCs metrics were recorded for models subsequently implemented on re-sampled data using the 4 re-sampling techniques: None (red), ROS (green), RUS (blue), and SMOTE (purple). Performances were aggregated using means across each chromosome, with error bars representing 1 standard deviation away from the mean.

**Supplementary Figure 4. Predictive importance is outweighed by transcription factor binding sites.** Stacked barplots illustrating the proportion of total predictive importance in models using optimal data level characteristics determined by the ensemble framework, for each resolution on K562. Proportions are separated by the set of annotations included in the feature space including chromatin state segmentation, histone modification, and transcription factor binding sites.

>>>>

**Supplementary Figure 1. Performing $log_{2}$ transformations normalized skewed distance densities across resolution for GM12878.** When measured in basepairs, distances were highly right skewed for all resolutions: (A) 10 kb, (B) 25 kb, (C) 50 kb, and (D) 100 kb. After $log_{2}$ transforming the distances, densities became much more gaussian distributed for all resolutions: (E) 10 kb, (F) 25 kb, (G) 50 kb, and (H) 100 kb. Each density curve represents a cell line specific annotation, and its relationship with all resolution specific genomic bins.

**Supplementary Figure 2. Performing $log_{2}$ transformations normalized skewed distance densities across resolution for K562.** When measured in basepairs, distances were highly right skewed for all resolutions: (A) 10 kb, (B) 25 kb, (C) 50 kb, and (D) 100 kb. After $log_{2}$ transforming the distances, densities became much more gaussian distributed for all resolutions: (E) 10 kb, (F) 25 kb, (G) 50 kb, and (H) 100 kb. Each density curve represents a cell line specific annotation, and its relationship with all resolution specific genomic bins.

@@

**Supplementary Figure 3. SMOTE re-sampled models built on distance-type predictors yielded optimal MCCs across resolutions for K562.** Performances of random forest classification models built on each of the 3 predictor types: (A) OC, (B) OP, and (C) Distance. Within each predictor type, and across resolution, training data was re-sampled using different techniques including: None (red), ROS (green), RUS (blue), and SMOTE (purple). Performances were aggregated using means across each chromosome, with error bars representing 1 standard deviation away from the mean.

**Supplementary Figure 4. The proportional importance of optimally built predictive models was outweighed by transcription factor binding sites across resolutions for K562.** Stacked barplots illustrate the proportion of total predictive importance in models using optimal data level characteristics determined by the ensemble framework (SMOTE re-sampling; with distance-type predictors), for each resolution on GM12878. Proportions are grouped by the class of cell line specific annotations in the feature space including chromatin state segmentations (CSS), histone modifications (HM), and transcription factor binding sites (TFBS).

### 2019-05-29 Manuscript/manuscript.Rmd
edits to figure legends

**Methods.** To investigate the problem of spatial genomic relationships and class imbalance problem, we developed an ensemble framework for optimizing prediction of TAD boundaries. We utilized several resampling techniques (random over- and undersampling, Synthetic Minority Over-sampling TEchnique (SMOTE)) coupled with various metrics for classifier performance evaluation (Area Under Receiver Operating Characteristics (AUCROC) curve, Precision-Recall curve (AUPRC), F1, Matthew Correlation Coefficient (MCC), among others) to select the best predictive model in class imbalance settings.

@@

A better threshold-free measure to use when dealing with imbalanced data is the area under the precision-recall curve (AUPRC).  The interpretation of the AUPRC of a model is similar to that of the AUROC. However, where as the baseline model performance according to the AUROC is given by 0.50, the baseline model performance according to the AUPRC is given by the ratio of true positives over the total number of cases. This enables a more realistic benchmark to evaluate a model on as this ratio will be lower than 0.5 for a test data set with imbalanced classes [@saito2015precision]. Another threshold-specific measure to consider is a composite metric involving all quadrants of the confusion matirix, such as Matthew's Correlation Coefficient (MCC) [@jeni2013facing; @chicco2017ten]. The MCC is, in essence, the correlation coefficient between the observed and predicted binary classifications [@brodersen2010balanced]. A value of positive 1 represents a perfect prediction, while a value of negative 1 represents total disagreement between prediction and observation. A value of 0 indicates no better than random prediction.

@@

Using the sets of genomic annotations, we created three types of predictors that were used as features for subsequent models to analyze their respective spatial relationship with TAD boundaries. They are described in detail below:

@@

3. **Log2 Distance**: For each genomic bin, the $log_{2}$ of the distance in bases (refered to hereafter simply as $\textit{distance}$), from the center of the bin to the center of the nearest region defined by a specific genomic annotation was calculated. The log transformation was used as a normalization technique in order to account for the skewness present in the distance densities <!--(Supplementary Figures 1)-->.

@@

The performance of each ensemble system in the framework was compared using the composite metric, Matthew's Correlation Coefficient. The MCC is defined as follows:

@@

We then went on the evaluate the effect of class imbalance on model performance. In order to set a baseline in performance, we first evaluated the least granular model, that is one implemented using OC predictor types, without data re-sampling. We found that models built on data with fully imbalanced classes exhibited poor MCCs across all resolutions (Figure 5A). We gradually increased the granularity by assessing models built on OP predictor types, while still not performing any data resampling. Here, the results were similar to that of OC types (Figure 5B). However, when including distance-type predictors, models were found to exhibit a marked improvement in MCC compared to the other two types (Figure 5C). Results were found to be consistently poor across resolutions and predictor types, when implemented on the K562 cell line (Supplementary Figure 3). Furthermore, we found that conventional metrics of performance (i.e. accuracy, sensitivity and specificity, and AUC) were less sensitive for models built on imbalanced data, regardless of cell line, resolution, or predictor-type (Supplementary Tables 6-7). [??? Need to know why PRCs give different results.]

@@

When assessing performances for models built on balanced data, we found that SMOTE re-sampling provided the optimal MCC consistently across both predictor type and resolution. Specifically, the most improvement in model performance was seen when combining SMOTE re-sampling with distance-type predictors. This was also the optimal combination for the K562 cell line (Supplementary Figure 3). These results imply that a best performing model, one that is able to best discriminate between genomic bins that are near versus far away from TAD boundaries, is one that is built on 10 kb resolution data, with distance-type predictors, and incorporates both random under- and over-sampling. Moreover, there was a clear degredation in model performance as data resolution decreased. Likewise, performances exhibited much more variability across chromosomes when built on lower resolutions.

@@

The experiments derived from our ensemble framework provide a useful guide toward the exploration of chromatin organization, while highlighting the impact of both class imbalance and feature engineering in TAD boundary prediction. Using the framework we demonstrated that random forest models built on imbalanced data exhibited poor performance, regardless of cell line, resolution, or predictor type. In our framework, we opted to leave the test set as a completely untouched holdout set, only performing the re-sampling, and subsequent learning, on the training set. This was done so as to incur the least amount of bias and so that all conclusions could be made soley between model characteristics. As a result, we showed that Matthew's Correlation Coefficient was much more sensitive to the imbalance present in the test set, compared to more popular metrics like AUROC, which attain artificially high values by correctly predicting majority class samples. When comparing re-sampling techniques, we found that SMOTE re-sampling outperformed models built on both simple ROS or RUS, implying that, in fact, a combination of both random under-sampling and random over-sampling is preferred in TAD boundary prediction. Furthermore, the most improvement was seen when combining SMOTE resampling on distance-type features when building random forest classifiers. These two characteristics exhibited consistenly better performance results across both the GM12878 and K562 cell lines. This confirmed our hypothesis that the most granular and spatially informative representation of the relationship between TAD boundaries and annotations resulted in the best performing model.

>>>>

**Methods.** To investigate the problem of spatial genomic relationships and class imbalance problem, we developed an ensemble framework for optimizing prediction of TAD boundaries. We utilized several resampling techniques (random over- and undersampling, Synthetic Minority Over-sampling TEchnique (SMOTE)) coupled with various metrics for classifier performance evaluation (Area Under Receiver Operating Characteristics (AUCROC) curve, Precision-Recall curve (AUPRC), F1, Matthews Correlation Coefficient (MCC), among others) to select the best predictive model in class imbalance settings.

@@

A better threshold-free measure to use when dealing with imbalanced data is the area under the precision-recall curve (AUPRC).  The interpretation of the AUPRC of a model is similar to that of the AUROC. However, where as the baseline model performance according to the AUROC is given by 0.50, the baseline model performance according to the AUPRC is given by the ratio of true positives over the total number of cases. This enables a more realistic benchmark to evaluate a model on as this ratio will be lower than 0.5 for a test data set with imbalanced classes [@saito2015precision]. Another threshold-specific measure to consider is a composite metric involving all quadrants of the confusion matirix, such as Matthews Correlation Coefficient (MCC) [@jeni2013facing; @chicco2017ten]. The MCC is, in essence, the correlation coefficient between the observed and predicted binary classifications [@brodersen2010balanced]. A value of positive 1 represents a perfect prediction, while a value of negative 1 represents total disagreement between prediction and observation. A value of 0 indicates no better than random prediction.

@@

Using the sets of cell line specific genomic annotations, we created three types of predictors that were used as features for subsequent models to analyze their respective spatial relationship with TAD boundaries. They are described in detail below:

@@

3. **Log2 Distance**: For each genomic bin, the $log_{2}$ of the distance in bases (refered to hereafter simply as $\textit{distance}$), from the center of the bin to the center of the nearest region defined by a specific genomic annotation was calculated. The log transformation was used as a normalization technique in order to account for the skewness present in the distance densities (Supplementary Figures 1-2).

@@

The performance of each ensemble system in the framework was compared using the composite metric, Matthews Correlation Coefficient. The MCC is defined as follows:

@@

We then went on the evaluate the effect of class imbalance on model performance. In order to set a baseline in performance, we first evaluated the least granular model, that is one implemented using OC predictor types, without data re-sampling. We found that models built on data with fully imbalanced classes exhibited poor MCCs across all resolutions (Figure 4A). We gradually increased the granularity by assessing models built on OP predictor types, while still not performing any data resampling. Here, the results were similar to that of OC types (Figure 4B). However, when including distance-type predictors, models were found to exhibit a marked improvement in MCC compared to the other two types (Figure 4C). Results were found to be consistently poor across resolutions and predictor types, when implemented on the K562 cell line (Supplementary Figure 3). Furthermore, we found that conventional metrics of performance (i.e. accuracy, sensitivity and specificity, and AUC) were less sensitive for models built on imbalanced data, regardless of cell line, resolution, or predictor-type (Supplementary Tables 6-7). [??? Need to know why PRCs give different results.]

@@

When assessing performances for models built on balanced data, we found that SMOTE re-sampling provided the optimal MCC consistently across both predictor type and resolution (Figure 4). Specifically, the most improvement in model performance was seen when combining SMOTE re-sampling with distance-type predictors. This was also the optimal combination for the K562 cell line (Supplementary Figure 3). These results imply that a best performing model, one that is able to best discriminate between genomic bins that are near versus far away from TAD boundaries, is one that is built on 10 kb resolution data, with distance-type predictors, and incorporates both random under- and over-sampling. Moreover, there was a clear degredation in model performance as data resolution decreased. Likewise, performances exhibited much more variability across chromosomes when built on lower resolutions.

@@

The experiments derived from our ensemble framework provide a useful guide toward the exploration of chromatin organization, while highlighting the impact of both class imbalance and feature engineering in TAD boundary prediction. Using the framework we demonstrated that random forest models built on imbalanced data exhibited poor performance, regardless of cell line, resolution, or predictor type. In our framework, we opted to leave the test set as a completely untouched holdout set, only performing the re-sampling, and subsequent learning, on the training set. This was done so as to incur the least amount of bias and so that all conclusions could be made soley between model characteristics. As a result, we showed that Matthews Correlation Coefficient was much more sensitive to the imbalance present in the test set, compared to more popular metrics like AUROC, which attain artificially high values by correctly predicting majority class samples. When comparing re-sampling techniques, we found that SMOTE re-sampling outperformed models built on both simple ROS or RUS, implying that, in fact, a combination of both random under-sampling and random over-sampling is preferred in TAD boundary prediction. Furthermore, the most improvement was seen when combining SMOTE resampling on distance-type features when building random forest classifiers. These two characteristics exhibited consistenly better performance results across both the GM12878 and K562 cell lines. This confirmed our hypothesis that the most granular and spatially informative representation of the relationship between TAD boundaries and annotations resulted in the best performing model.
